---
title: "GSE19080"
output: html_document
---

Non-Infected 8 vs Adult T-cell leukemia

# library
```{r}
library(GEOquery)
library(limma)
library(Biobase)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(tidyr)
library(purrr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ReactomePA)
library(DOSE)
library(enrichR)
library(tibble)
```

```{r}
gse <- getGEO("GSE19080", GSEMatrix = TRUE)
```

```{r}
# expression matrix
exprs_data <- exprs(gse[[1]])

# sample metadata (phenodata)
pheno_data <- pData(gse[[1]])

# feature data (annotation)
feature_data <- fData(gse[[1]])

head(exprs_data)
head(pheno_data)
head(feature_data)
```

```{r}
annot <- gse[[1]]@featureData@data
exprs_with_symbol <- cbind(SYMBOL = annot$SYMBOL, exprs_data)
head(exprs_with_symbol)
```

```{r}
group <- factor(pheno_data[["disease state:ch1"]])
group <- as.character(group)
group <- dplyr::case_when(
  grepl("Adult T-cell leukemia", group) ~ "ATL",
  grepl("Asymtomatic Carrier", group) ~ "AC",  
  grepl("Healthy Donor", group) ~ "NI",
  grepl("HAM/TSP", group) ~ "HAM_TSP",         
  TRUE ~ group
)
group <- factor(group, levels = c("NI", "AC", "HAM_TSP", "ATL"))

```

# 1. Preprocessing (Boxplot, PCA)
```{r}
# boxplot before vs after
# Log2
norm_exprs_data <- log2(exprs_data) 

boxplot(exprs_data[, 1:38], main = "After Normalization", las = 2)
boxplot(norm_exprs_data[, 1:38], main = "After Normalization", las = 2)
norm_exprs_data <- exprs_data
```

```{r}
selected_idx <- group %in% c("NI", "AC", "HAM/TSP")

exprs_filtered <- exprs_data[, selected_idx]
group_filtered <- droplevels(group[selected_idx])

t_expr <- t(exprs_filtered)

pca_res <- prcomp(t_expr, center = TRUE, scale. = TRUE)

pca_df <- as.data.frame(pca_res$x[, 1:2])  # PC1, PC2
pca_df$Group <- group_filtered

library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of NI vs AC vs HAM/TSP Samples", x = "PC1", y = "PC2") +
  theme(plot.title = element_text(hjust = 0.5))
```


# 2. DEG (lo2fc 1.5, adjusted p-value 0.05) volcano plot, heatmap
```{r}
# 
selected_idx <- group %in% c("HAM_TSP", "NI")
exprs_data <- exprs_data[, selected_idx]
group <- droplevels(group[selected_idx])

# design
design <- model.matrix(~0 + group)  # delete intercept
colnames(design) <- levels(group)   #

# linear model fitting
fit <- lmFit(exprs_data, design)

# contrast (ATL vs Control)
contrast_matrix <- makeContrasts(HAM_TSPvsNI = HAM_TSP - NI, levels = design)

# apply contrast & eBayes
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

# substract DEG
deg_results <- topTable(fit2, coef = "HAM_TSPvsNI", adjust = "BH", number = Inf)
deg_results <- deg_results %>% rownames_to_column(var = "ID")
deg_results$ID <- as.character(deg_results$ID)
annot$ID <- as.character(annot$ID)
deg_results <- left_join(deg_results, annot, by = "ID")
deg_results <- dplyr::select(deg_results, SYMBOL, everything())

# filtering: |log2FC| > 1.5, adj.P.Val < 0.05
deg_signif <- deg_results %>%
  dplyr::filter(abs(logFC) > 1.5, adj.P.Val < 0.05)

```


```{r}
# attatch gene symbol
deg_signif$GENE_SYMBOL <- feature_data$GENE_SYMBOL[match(rownames(deg_signif), rownames(feature_data))]
deg_results$GENE_SYMBOL <- feature_data$GENE_SYMBOL[match(rownames(deg_results), rownames(feature_data))]
head(deg_signif)
```

# Volcano plot
```{r}
deg_results$threshold <- with(deg_results, ifelse(adj.P.Val < 0.05 & abs(logFC) > 1.5,
                                                  ifelse(logFC > 1.5, "Up", "Down"), "NS"))

ggplot(deg_results, aes(x = logFC, y = -log10(adj.P.Val), color = threshold)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "grey")) +
  theme_minimal() +
  labs(title = "Volcano Plot: ATL vs Control",
       x = "log2 Fold Change", y = "-log10 Adjusted P-value")
```

```{r}
top_genes <- head(rownames(deg_signif), 50)

heat_data <- exprs_data[top_genes, ]

heat_data_scaled <- t(scale(t(heat_data)))

probe_to_symbol <- feature_data$SYMBOL
names(probe_to_symbol) <- rownames(feature_data)

gene_symbols <- probe_to_symbol[top_genes]

gene_symbols_clean <- ifelse(is.na(gene_symbols) | gene_symbols == "",
                             top_genes, gene_symbols)
gene_symbols_clean <- make.unique(gene_symbols_clean)

rownames(heat_data_scaled) <- gene_symbols_clean

annotation_col <- data.frame(Group = group)
rownames(annotation_col) <- colnames(heat_data_scaled)

pheatmap(heat_data_scaled,
         annotation_col = annotation_col,
         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         main = "Top 50 DEGs (Gene Symbols)")

```

# 3. GSEA analysis of DEG of GSE33615 reactome analysis / KEGG analysis
Gene Set Enrichment Analysis (logFC)
Pathway
```{r}
# logFC
gene_list <- deg_results$logFC
names(gene_list) <- deg_results$SYMBOL
gene_list <- sort(gene_list, decreasing = TRUE)
symbol2entrez <- bitr(names(gene_list), fromType = "SYMBOL",
                      toType = "ENTREZID",
                      OrgDb = org.Hs.eg.db)

# geneList (ENTREZID)
gene_list_entrez <- gene_list[symbol2entrez$SYMBOL]
names(gene_list_entrez) <- symbol2entrez$ENTREZID
gene_list_entrez <- sort(gene_list_entrez, decreasing = TRUE)

```

# KEGG
```{r}
gsea_kegg <- gseKEGG(geneList = gene_list_entrez,
                     organism = "hsa",
                     pvalueCutoff = 0.05,
                     verbose = FALSE)
dotplot(gsea_kegg, showCategory = 20)
```

# GO Biological Process
```{r}
gsea_go <- gseGO(geneList = gene_list_entrez,
                 OrgDb = org.Hs.eg.db,
                 ont = "BP",
                 keyType = "ENTREZID",
                 pvalueCutoff = 0.05,
                 verbose = FALSE)
dotplot(gsea_go, showCategory = 20)
```

# Reactome
```{r}
gsea_reactome <- gsePathway(geneList = gene_list_entrez,
                            organism = "human",
                            pvalueCutoff = 0.05,
                            verbose = FALSE)
dotplot(gsea_reactome, showCategory = 20)

```

# ORA (Over-Representation Analysis)
Whether DEG is concentrated in specific pathways
KEGG: gene - pathway - disease
Reactome: modeling specific molecule flow of biological reaction in cell

# KEGG ORA
```{r}
# extract DEG lists
deg_genes <- deg_results %>%
  filter(adj.P.Val < 0.05 & abs(logFC) > 1.5) %>%
  filter(!is.na(SYMBOL) & SYMBOL != "")

# SYMBOL → ENTREZID
deg_entrez <- bitr(deg_genes$SYMBOL, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# enrichKEGG
enrich_kegg <- enrichKEGG(gene = deg_entrez$ENTREZID,
                          organism = "hsa",
                          pvalueCutoff = 0.05)

dotplot(enrich_kegg, showCategory = 30)

```

# Reactome ORA
```{r}
enrich_reactome <- enrichPathway(gene = deg_entrez$ENTREZID,
                                 organism = "human",
                                 pvalueCutoff = 0.05,
                                 readable = TRUE)

# Visualization
dotplot(enrich_reactome, showCategory = 30)
```

# 4. Pontential Pharmacology of Identified Targets DEGs
```{r}
stat_genes <- c("MLL", "STAT1", "STAT3", "STAT5A", "STAT5B", "IRF4", "JAK3", "IL2RA")

stat_probes <- feature_data %>%
  filter(SYMBOL %in% stat_genes) %>%
  dplyr::select(ID, SYMBOL)

stat_exprs <- exprs_data[rownames(exprs_data) %in% stat_probes$ID, ]

rownames(stat_exprs) <- stat_probes$SYMBOL[match(rownames(stat_exprs), stat_probes$ID)]

```

```{r}
library(reshape2)
stat_exprs_df <- as.data.frame(stat_exprs)
stat_exprs_df$GENE <- rownames(stat_exprs_df)
long_df <- melt(stat_exprs_df, id.vars = "GENE", variable.name = "Sample", value.name = "Expression")

long_df$Group <- group[match(long_df$Sample, colnames(exprs_data))]

# ggplot boxplot
library(ggplot2)
ggplot(long_df, aes(x = Group, y = Expression, fill = Group)) +
  geom_boxplot() +
  facet_wrap(~GENE, scales = "free_y") +
  theme_minimal() +
  labs(title = "STAT Genes Expression: ATL vs Control")

long_df %>%
  group_by(GENE) %>%
  summarise(p_value = wilcox.test(Expression ~ Group)$p.value)
```
