---
title: "GSE33615"
output: html_document
---

# library
```{r}
library(GEOquery)
library(limma)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(tibble)
library(dplyr)
library(enrichplot)
library(ReactomePA)
library(clusterProfiler)
library(org.Hs.eg.db)
```


# Data preparation
```{r}
# GEO ID
geo_id <- "GSE33615"

# data download
gse <- getGEO(geo_id, GSEMatrix = TRUE)
```

```{r}
exprSet <- gse[[1]]

# (expression values)
exprs_data <- exprs(exprSet)

# metatdata
pheno_data <- pData(exprSet)

# annotation (probe information)
feature_data <- fData(exprSet)

# ATL vs Control
source_labels <- trimws(as.character(pheno_data$source_name_ch1))

group <- factor(ifelse(grepl("^ATL", source_labels), "ATL", "NI"))
```


```{r}
# Acute Chronic Uninfected
title_labels <- trimws(as.character(pheno_data$characteristics_ch1.1))

group <- case_when(
  grepl("Acute", title_labels) ~ "Acute",
  grepl("Chronic", title_labels) ~ "Chronic",
  grepl("uninfected", title_labels) ~ "Control",
  TRUE ~ "Other"
)

group <- factor(group)
table(group)
#
```

```{r}
annot <- gse[[1]]@featureData@data
```


# Preprocessing Normalization
```{r}


# normalizeBetweenArrays
norm_exprs_data <- normalizeBetweenArrays(exprs_data, method = "quantile")

# Log2
norm_exprs_data <- log2(norm_exprs_data) 

# boxplot before vs after
boxplot(exprs_data[, 1:73], main = "Before Normalization", las = 2)
boxplot(norm_exprs_data[, 1:73], main = "After Normalization", las = 2)
```

```{r}
# Select ATL Control
selected_idx <- group %in% c("NI", "ATL")

exprs_filtered <- norm_exprs_data[, selected_idx]
group_filtered <- droplevels(group[selected_idx])

t_expr <- t(exprs_filtered)

# perform PCA
pca_res <- prcomp(t_expr, center = TRUE, scale. = TRUE)

# result frame
pca_df <- as.data.frame(pca_res$x[, 1:2])  # PC1, PC2
pca_df$Group <- group_filtered

# visualization
library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "ATL vs Control", x = "PC1", y = "PC2") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
t_expr_all <- t(norm_exprs_data)
group_all <- group  

pca_res_all <- prcomp(t_expr_all, center = TRUE, scale. = TRUE)

pca_df_all <- as.data.frame(pca_res_all$x[, 1:2])
pca_df_all$Group <- group_all

ggplot(pca_df_all, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of All Groups", x = "PC1", y = "PC2") +
  theme(plot.title = element_text(hjust = 0.5))

```


# DEG (lo2fc 1.5, adjusted p-value 0.05)
```{r}
# design
design <- model.matrix(~0 + group)  # delete intercept
colnames(design) <- levels(group)   #

# linear model fitting
fit <- lmFit(norm_exprs_data, design)

# contrast (ATL vs Control)
contrast_matrix <- makeContrasts(ATLvsNI = ATL - NI, levels = design)

# apply contrast & eBayes
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

# substract DEG
deg_results <- topTable(fit2, coef = "ATLvsNI", adjust = "BH", number = Inf)
deg_results <- deg_results %>% rownames_to_column(var = "ID")
deg_results$ID <- as.character(deg_results$ID)
annot$ID <- as.character(annot$ID)
deg_results <- left_join(deg_results, annot, by = "ID")
deg_results <- dplyr::select(deg_results, GENE_SYMBOL, everything())

# delete NA in GENE_SYMBOL
deg_results <- deg_results %>%
  filter(!is.na(GENE_SYMBOL) & GENE_SYMBOL != "")
```


```{r}
# logFC probe
deg_results <- deg_results %>%
  group_by(GENE_SYMBOL) %>%
  slice_max(order_by = abs(logFC), n = 1, with_ties = FALSE) %>%
  ungroup()

# filtering: |log2FC| > 1.5, adj.P.Val < 0.05
deg_signif <- deg_results %>%
  dplyr::filter(abs(logFC) > 1.5, adj.P.Val < 0.05)
```

# Volcano plot
```{r}
# Up: 800 Down 770
deg_results$threshold <- with(deg_results, ifelse(adj.P.Val < 0.05 & abs(logFC) > 1.5,
                                                  ifelse(logFC > 1.5, "Up", "Down"), "NS"))

ggplot(deg_results, aes(x = logFC, y = -log10(adj.P.Val), color = threshold)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "grey")) +
  theme_minimal() +
  labs(title = "ATL vs Control",
       x = "log2 Fold Change", y = "-log10 Adjusted P-value")
```

# 3. GSEA analysis of DEG of GSE33615 reactome analysis / KEGG analysis
Gene Set Enrichment Analysis (logFC)
Pathway
```{r}
# logFC
gene_list <- deg_results$logFC
names(gene_list) <- deg_results$GENE_SYMBOL
gene_list <- sort(gene_list, decreasing = TRUE)
symbol2entrez <- bitr(names(gene_list), fromType = "SYMBOL",
                      toType = "ENTREZID",
                      OrgDb = org.Hs.eg.db)

# geneList (ENTREZID)
gene_list_entrez <- gene_list[symbol2entrez$SYMBOL]
names(gene_list_entrez) <- symbol2entrez$ENTREZID
gene_list_entrez <- sort(gene_list_entrez, decreasing = TRUE)

```

# KEGG
```{r}
gsea_kegg <- gseKEGG(geneList = gene_list_entrez,
                     organism = "hsa",
                     pvalueCutoff = 0.05,
                     verbose = FALSE)
dotplot(gsea_kegg, showCategory = 20)
```

# GO Biological Process
```{r}
gsea_go <- gseGO(geneList = gene_list_entrez,
                 OrgDb = org.Hs.eg.db,
                 ont = "BP",
                 keyType = "ENTREZID",
                 pvalueCutoff = 0.05,
                 verbose = FALSE)
dotplot(gsea_go, showCategory = 20)
```

# Reactome
```{r}
gsea_reactome <- gsePathway(geneList = gene_list_entrez,
                            organism = "human",
                            pvalueCutoff = 0.05,
                            verbose = FALSE)
dotplot(gsea_reactome, showCategory = 20)

```

# ORA (Over-Representation Analysis)
Whether DEG is concentrated in specific pathways
KEGG: gene - pathway - disease
Reactome: modeling specific molecule flow of biological reaction in cell

# KEGG ORA
```{r}
# extract DEG lists
deg_genes <- deg_results %>%
  filter(adj.P.Val < 0.05 & abs(logFC) > 1.5) %>%
  filter(!is.na(GENE_SYMBOL) & GENE_SYMBOL != "")

# SYMBOL → ENTREZID
deg_entrez <- bitr(deg_genes$GENE_SYMBOL, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# enrichKEGG
enrich_kegg <- enrichKEGG(gene = deg_entrez$ENTREZID,
                          organism = "hsa",
                          pvalueCutoff = 0.05)

dotplot(enrich_kegg, showCategory = 30)

```

# Reactome ORA
```{r}
enrich_reactome <- enrichPathway(gene = deg_entrez$ENTREZID,
                                 organism = "human",
                                 pvalueCutoff = 0.05,
                                 readable = TRUE)

# Visualization
dotplot(enrich_reactome, showCategory = 23)
```

# JAK-STAT signaling pathway genes
```{r}
# result table
enrich_res <- enrich_kegg@result

# JAK-STAT pathway
jak_row <- enrich_res[grep("JAK-STAT signaling pathway", enrich_res$Description), ]

# gene list ENTREZ to SYMBOL
jak_genes <- strsplit(jak_row$geneID, "/")[[1]]

jak_gene_symbols <- bitr(jak_genes,
                         fromType = "ENTREZID",
                         toType = "SYMBOL",
                         OrgDb = org.Hs.eg.db)

jak_gene_symbols$SYMBOL
```

```{r}
# Cytokine-cytokine receptor
cyto_row <- enrich_res[grep("Cytokine-cytokine receptor", enrich_res$Description), ]

# gene list ENTREZ to SYMBOL
cyto_genes <- strsplit(cyto_row$geneID, "/")[[1]]

cyto_gene_symbols <- bitr(cyto_genes,
                         fromType = "ENTREZID",
                         toType = "SYMBOL",
                         OrgDb = org.Hs.eg.db)

cyto_gene_symbols$SYMBOL
```

```{r}
intersect_genes <- intersect(jak_gene_symbols$SYMBOL, cyto_gene_symbols$SYMBOL)
intersect_genes
```

```{r}
stat_genes <- c("MLL", "STAT1", "STAT3", "STAT5A", "STAT5B", "IRF4", "JAK3", "IL2RA", "MYB")

stat_probes <- feature_data %>%
  filter(GENE_SYMBOL %in% stat_genes) %>%
  dplyr::select(ID, GENE_SYMBOL)

stat_exprs <- exprs_data[rownames(exprs_data) %in% stat_probes$ID, ]

rownames(stat_exprs) <- stat_probes$GENE_SYMBOL[match(rownames(stat_exprs), stat_probes$ID)]

```

```{r}
library(reshape2)
stat_exprs_df <- as.data.frame(stat_exprs)
stat_exprs_df$GENE <- rownames(stat_exprs_df)
long_df <- melt(stat_exprs_df, id.vars = "GENE", variable.name = "Sample", value.name = "Expression")

long_df$Group <- group[match(long_df$Sample, colnames(exprs_data))]

# ggplot boxplot
library(ggplot2)
ggplot(long_df, aes(x = Group, y = Expression, fill = Group)) +
  geom_boxplot() +
  facet_wrap(~GENE, scales = "free_y") +
  theme_minimal() +
  labs(title = "STAT Genes Expression: ATL vs Control")

long_df %>%
  group_by(GENE) %>%
  summarise(p_value = wilcox.test(Expression ~ Group)$p.value)
```

```{r}
stat_genes <- c("MLL", "STAT1", "STAT3", "STAT5A", "STAT5B", "IRF4", "JAK3", "IL2RA")

stat_probes <- feature_data %>%
  filter(GENE_SYMBOL %in% stat_genes) %>%
  dplyr::select(ID, GENE_SYMBOL)

stat_exprs <- exprs_data[rownames(exprs_data) %in% stat_probes$ID, ]

rownames(stat_exprs) <- stat_probes$GENE_SYMBOL[match(rownames(stat_exprs), stat_probes$ID)]

```

```{r}
library(reshape2)
stat_exprs_df <- as.data.frame(stat_exprs)
stat_exprs_df$GENE <- rownames(stat_exprs_df)
long_df <- melt(stat_exprs_df, id.vars = "GENE", variable.name = "Sample", value.name = "Expression")

long_df$Group <- group[match(long_df$Sample, colnames(exprs_data))]

# ggplot boxplot
library(ggplot2)
ggplot(long_df, aes(x = Group, y = Expression, fill = Group)) +
  geom_boxplot() +
  facet_wrap(~GENE, scales = "free_y") +
  theme_minimal() +
  labs(title = "STAT Genes Expression: ATL vs Control")

long_df %>%
  group_by(GENE) %>%
  summarise(p_value = wilcox.test(Expression ~ Group)$p.value)
```